On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-09-24 paper.tex
Rewritten proof of \ref{theo: probmax, sq} to clarify symmetry of considered tiles

A square grid has $a=b$. For real-values segment lengths it can be further assumed that $a=1$. For $a \neq 1$ the results to be obtained apply with $\len$ replaced by $\len/a$.

@@

Consider a length $\len$ with corresponding maximum number of tiles $\tiles = \funt(\len)$. The first endpoint of the segment is assumed to be in the reference tile with lower-left corner $(0,0)$. The segment visits $\tiles$ tiles almost surely if and only if its second endpoint is in a tile with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, $i+j-1=\tiles$, or in a tile symmetrical to this in the other three quadrants of the plane. Figure~\ref{fig: probmax_possible_tiles} shows the reference tile, with solid line, and some of the tiles that fulfil this condition, with dashed line, for $\tiles=5$. The desired probability can be computed by considering only the tiles in the first quadrant and multiplying by $4$.

% ***Does this figure add something, or is it better to remove it?
\begin{figure}
\centering%
\includegraphics[width=.52\textwidth]{probmax_possible_tiles}%
\caption{Tiles in which the second endpoint can be for a given number of visited tiles $\tiles$ in a square grid (example for $\tiles=5$)
}%
\label{fig: probmax_possible_tiles}%
\end{figure}%

For any $i,j$ with $i+j-1=\tiles$ and for a given orientation $\theta$, the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ if and only if shifting the segment so that its second endpoint has coordinates $(i-1,j-1)$ results in the first endpoint being in the reference tile. It cannot be the case that the segment ``overshoots'' past this tile, because that would result in a number of visited tiles exceeding the assumed maximum $\tiles$.

@@

There is a range of possible values of $\theta$ for which the above condition is satisfied. This range corresponds to the part of the arc with solid line in Figure~\ref{fig: probmax_funt_odd}. For convenience, the angle $\theta$ is defined clockwise from the downward direction. For each $\theta$ in the allowed range, the valid positions for the first endpoint of the (shifted) segment are in the shaded rectangle, whose dimensions depend on $\theta$. The area occupied by this rectangle is the probability $\probmax_{i-1,j-1}(\len,\theta)$ that the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ conditioned on $\theta$. The desired probability, $\probmax(\len)$, is obtained by averaging $\probmax_{i-1,j-1}(\len, \theta)$ over $\theta$ and summing over all $i-1,j-1$ that can be reached from the reference tile.

There are two situations depending on whether $\tiles$ is odd or even. Consider $\tiles$ odd first. Under this assumption, the range of lengths corresponding to $\tiles$ is given by \eqref{eq: square, odd max tiles, lengths}. For any length in this interval, there are some positions within the reference tile from which the tile with lower-left coordinate $((\tiles-1)/2, (\tiles-1)/2)$ can be reached. Equivalently, the shifted segment with one endpoint at $((\tiles-1)/2, (\tiles-1)/2)$ can reach the reference tile. This is illustrated in Figure~\ref{fig: probmax_funt_odd_diag}. In addition, if 
\begin{equation}
\label{eq: probmax len odd2}
\len > \frac{\sqrt{(\tiles-5)^2 + (\tiles-1)^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $((\tiles-3)/2, (\tiles+1)/2)$ and $((\tiles+1)/2, (\tiles-3)/2)$ are also reachable from the reference tile; or equivalently the shifted segment can reach the reference tile, as shown in Figure~\ref{fig: probmax_funt_odd_offdiag}. This can only happen for $\tiles \geq 5$ (for $\tiles=3$ the inequality \eqref{eq: probmax len odd2} cannot be satisfied for lengths in the range given by \eqref{eq: square, odd max tiles, lengths}). In this case, by symmetry, each of those two tiles contributes the same probability. Other tiles corresponding to the same $\tiles$ need not be considered, because they cannot be reached with lengths satisfying \eqref{eq: square, odd max tiles, lengths}.

For $\tiles$ even, which corresponds to lengths in the interval \eqref{eq: square, even max tiles, lengths}, the two tiles with lower-left coordinates $(\tiles/2-1, \tiles/2)$ and $(\tiles/2, \tiles/2-1)$ can always be reached from some positions within the reference tile. In addition, if
\begin{equation}
\label{eq: probmax len even2}
\len > \frac{\sqrt{((\tiles-6)^2 + \tiles^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $(\tiles/2-2, \tiles/2+1)$ and $(\tiles/2+1, \tiles/2-2)$ have to be considered too. This can only happen for $\tiles \geq 8$. No other tiles can be reached from the reference tile.

Thus, two cases need to be distinguished for $\tiles$ odd, depending on whether \eqref{eq: probmax len odd2} is satisfied or not; and similarly two cases exist for $\tiles$ even, depending on whether \eqref{eq: probmax len even2} holds or not. These are, respectively, the four cases in \eqref{eq: probmax}.

In the first case, as has been argued, only the tile with lower-left coordinates $((\tiles-1)/2, (\tiles-1)/2)$ is reachable. For a given $\theta$, the conditional probability $\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)$ is the area of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_diag}:
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta) = \left(\len\sin\theta - \frac{\tiles-3}2\right) \left(\len\cos\theta - \frac{\tiles-3}2\right).
\end{equation}
The range of allowed values for $\theta$, as can be seen from the figure, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{\tiles-3}{2\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{\tiles-3}{2\len}.
\end{align}
Computing the average of \eqref{eq: probmax: probmax theta odd} over $\theta$, multiplying by $4$ to include all quadrants and comparing with \eqref{eq: g def} gives $\probmax(\len)$ as
\begin{equation}
\label{eq: probmax: probmax odd}
\begin{split}
\probmax(\len) & = 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)}{2\pi} \diff \theta \\
& = \frac 2 \pi \int_{\arcsin\frac{\tiles-3}{2\len}}^{\arccos\frac{\tiles-3}{2\len}} \left( \len\sin\theta - \frac{\tiles-3} 2 \right)  \left( \len\cos\theta - \frac{\tiles-3} 2 \right) \diff \theta \\
& = g\left( \len, \tiles-3, \tiles-3 \right ),
\end{split}
\end{equation}
in accordance with \eqref{eq: probmax}.

For the second case in \eqref{eq: probmax}, $\probmax(\len)$ is obtained as in the preceding paragraph plus an additional term corresponding to the contribution of the tile with lower-left corner $((\tiles-3)/2, (\tiles+1)/2)$, multiplied by $2$ to account for its symmetrical tile $((\tiles+1)/2, (\tiles-3)/2)$ and by $4$ to include all quadrants. Observing the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_offdiag},
\begin{equation}
\label{eq: probmax: probmax theta odd2}
\probmax_{(\tiles-3)/2, (\tiles+1)/2}(\len, \theta) = \left(\len\sin\theta - \frac{\tiles-5}2\right) \left(\len\cos\theta - \frac{\tiles-1}2\right).
\end{equation}
The integration interval $(\theta_0, \theta_1)$ is in this case
\begin{align}
\theta_0 & = \arcsin\frac{\tiles-5}{2\len}, \\
\theta_1 & = \arccos\frac{\tiles-1}{2\len}.
\end{align}
The additional contribution to $\probmax(\len)$, given by
\begin{multline}
8 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(\tiles-3)/2, (\tiles+1)/2}(\len, \theta)}{2\pi} \diff \theta \\
= \frac 4 \pi \int_{\arcsin\frac{\tiles-5}{2\len}}^{\arccos\frac{\tiles-1}{2\len}} \left( \len\sin\theta - \frac{\tiles-5} 2 \right)  \left( \len\cos\theta - \frac{\tiles-1} 2 \right) \diff \theta,
\end{multline}
is recognized to be $2 g(\len, \tiles-5, \tiles-1)$. This proves the second equality in \eqref{eq: probmax}.

>>>>

A square grid has $a=b$. For real-valued segment lengths it can be further assumed that $a=1$. For $a \neq 1$ the results to be obtained apply with $\len$ replaced by $\len/a$.

@@

The first endpoint of the segment is assumed to be in the reference tile with lower-left corner $(0,0)$. For an arbitrary value $\tiles$, the segment visits $\tiles$ tiles almost surely if and only if its second endpoint is in a tile with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, $i+j-1=\tiles$, or in a tile symmetrical to this with respect to the line $x=1/2$ (with $i \leq 0$), or $y=1/2$ (with $j \leq 0$), or both.

Consider $i,j\geq 1$ for the moment, and let $\tiles = \funt(\len)$. For a given orientation $\theta$, the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ if and only if shifting the segment so that its second endpoint has coordinates $(i-1,j-1)$ results in the first endpoint being in the reference tile. It cannot be the case that the segment ``overshoots'' past this tile, because that would result in a number of visited tiles exceeding the assumed maximum, $\tiles$.

@@

There is a range of possible values of $\theta$ for which the above condition is satisfied. This range corresponds to the part of the arc with solid line in Figure~\ref{fig: probmax_funt_odd}. For convenience, the angle $\theta$ is defined clockwise from the downward direction. For each $\theta$ in the allowed range, the valid positions for the first endpoint of the (shifted) segment are in the shaded rectangle, whose dimensions depend on $\theta$. The area occupied by this rectangle is the probability $\probmax_{i-1,j-1}(\len,\theta)$ that the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ conditioned on $\theta$. The desired probability, $\probmax(\len)$, will be obtained by averaging $\probmax_{i-1,j-1}(\len, \theta)$ over $\theta$, summing over all $i, j \geq 1$ such that the tile with lower-left corner $(i-1, j-1)$ can be reached from the reference tile, and then using symmetry to take into account the tiles with $i \leq 0$ or $j \leq 0$.

The situation is different depending on whether $\tiles$ is odd or even. Consider $\tiles$ odd first. Under this assumption, the range of lengths corresponding to the condition $\tiles = \funt(\len)$ is given by \eqref{eq: square, odd max tiles, lengths}. For any length in this interval, there are some positions within the reference tile from which the tile with lower-left coordinate $((\tiles-1)/2, (\tiles-1)/2)$ can be reached. Equivalently, the shifted segment with one endpoint at $((\tiles-1)/2, (\tiles-1)/2)$ can reach the reference tile. This is illustrated in Figure~\ref{fig: probmax_funt_odd_diag}. In addition, if 
\begin{equation}
\label{eq: probmax len odd2}
\len > \frac{\sqrt{(\tiles-5)^2 + (\tiles-1)^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $((\tiles-3)/2, (\tiles+1)/2)$ and $((\tiles+1)/2, (\tiles-3)/2)$ are also reachable from the reference tile; or equivalently the shifted segment can reach the reference tile, as shown in Figure~\ref{fig: probmax_funt_odd_offdiag}. This can only happen for $\tiles \geq 5$ (for $\tiles=3$ the inequality \eqref{eq: probmax len odd2} cannot be satisfied for lengths in the range given by \eqref{eq: square, odd max tiles, lengths}). In this case, by symmetry, each of those two tiles contributes the same probability. Other tiles with $i,j\geq 1$ corresponding to the same $\tiles$ do not need to be considered, because they cannot be reached with lengths satisfying \eqref{eq: square, odd max tiles, lengths}.

For $\tiles$ even, which corresponds to lengths in the interval \eqref{eq: square, even max tiles, lengths}, the two tiles with lower-left coordinates $(\tiles/2-1, \tiles/2)$ and $(\tiles/2, \tiles/2-1)$ can always be reached from some positions within the reference tile. In addition, if
\begin{equation}
\label{eq: probmax len even2}
\len > \frac{\sqrt{((\tiles-6)^2 + \tiles^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $(\tiles/2-2, \tiles/2+1)$ and $(\tiles/2+1, \tiles/2-2)$ have to be considered too. This can only happen for $\tiles \geq 8$. No other tiles with $i,j\geq 1$ can be reached from the reference tile.

The above two paragraphs show that, among all possible tiles with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, those whose centers lie on the lines $x=1/2$ or $y=1/2$ (i.e.~those with $i=1$ of $j=1$) are never reachable from the reference tile, in the sense that a segment that touches the maximum number of tiles cannot have its second endpoint in those tiles. This implies that $\probmax(\len)$ can be computed by considering only the tiles with $i,j\geq 1$ (actually $i,j \geq 2$), and then multiplying by $4$ to include the symmetrical tiles with respect to $x=1/2$ and $y=1/2$. As an example, Figure~\ref{fig: probmax_possible_tiles} shows the reference tile (solid line) and the tiles that can contain the second endpoint (dotted lines) for $\funt(\len) = 5$.

\begin{figure}
\centering%
\includegraphics[width=.52\textwidth]{probmax_possible_tiles}%
\caption{Tiles that can contain the second endpoint of a segment that visits the maximum number of tiles $\funt(\len)$ in a square grid (example for $\funt(\len)=5$)%
}%
\label{fig: probmax_possible_tiles}%
\end{figure}%

According to the preceding analysis, two cases need to be distinguished for $\tiles$ odd, depending on whether \eqref{eq: probmax len odd2} is satisfied or not; and similarly two cases exist for $\tiles$ even, depending on whether \eqref{eq: probmax len even2} holds or not. These are, respectively, the four cases in \eqref{eq: probmax}.

In the first case indicated in \eqref{eq: probmax}, as argued, $\tiles$ is odd and only the tile with lower-left coordinates $((\tiles-1)/2, (\tiles-1)/2)$ is reachable. For a given $\theta$, the conditional probability $\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)$ is the area of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_diag}:
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta) = \left(\len\sin\theta - \frac{\tiles-3}2\right) \left(\len\cos\theta - \frac{\tiles-3}2\right).
\end{equation}
The range of allowed values for $\theta$, as can be seen from the figure, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{\tiles-3}{2\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{\tiles-3}{2\len}.
\end{align}
Computing the average of \eqref{eq: probmax: probmax theta odd} over $\theta$, multiplying by $4$ to include all quadrants and comparing with \eqref{eq: g def} gives $\probmax(\len)$ as
\begin{equation}
\label{eq: probmax: probmax odd}
\begin{split}
\probmax(\len) & = 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(\tiles-1)/2, (\tiles-1)/2}(\len, \theta)}{2\pi} \diff \theta \\
& = \frac 2 \pi \int_{\arcsin\frac{\tiles-3}{2\len}}^{\arccos\frac{\tiles-3}{2\len}} \left( \len\sin\theta - \frac{\tiles-3} 2 \right)  \left( \len\cos\theta - \frac{\tiles-3} 2 \right) \diff \theta \\
& = g\left( \len, \tiles-3, \tiles-3 \right ),
\end{split}
\end{equation}
in accordance with \eqref{eq: probmax}.

For the second case in \eqref{eq: probmax}, $\probmax(\len)$ is obtained as above plus an additional term corresponding to the tile with lower-left corner $((\tiles-3)/2, (\tiles+1)/2)$, multiplied by $2$ to account for its symmetrical tile about the line $y=x$, namely that with lower-left corner $((\tiles+1)/2, (\tiles-3)/2)$, and by $4$ to include all quadrants. Observing the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_offdiag},
\begin{equation}
\label{eq: probmax: probmax theta odd2}
\probmax_{(\tiles-3)/2, (\tiles+1)/2}(\len, \theta) = \left(\len\sin\theta - \frac{\tiles-5}2\right) \left(\len\cos\theta - \frac{\tiles-1}2\right).
\end{equation}
The integration interval $(\theta_0, \theta_1)$ is in this case
\begin{align}
\theta_0 & = \arcsin\frac{\tiles-5}{2\len}, \\
\theta_1 & = \arccos\frac{\tiles-1}{2\len}.
\end{align}
The additional contribution to $\probmax(\len)$, given by
\begin{multline}
8 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(\tiles-3)/2, (\tiles+1)/2}(\len, \theta)}{2\pi} \diff \theta \\
= \frac 4 \pi \int_{\arcsin\frac{\tiles-5}{2\len}}^{\arccos\frac{\tiles-1}{2\len}} \left( \len\sin\theta - \frac{\tiles-5} 2 \right)  \left( \len\cos\theta - \frac{\tiles-1} 2 \right) \diff \theta,
\end{multline}
is recognized to be $2 g(\len, \tiles-5, \tiles-1)$. This proves the second equality in \eqref{eq: probmax}.
